---
title: "Enrichment analysis for GO and KEGG Pathway"
author: "Kai Guo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This document describes how to use the richR package.
Below is a simple manual for using the package

## Introduction
You can do enrichment analysis for different type of annotation data (GO,KEGG,Reactome(may need reactome.db if you work with Human). If you had bioAnno installed, you can do enrichment with all types of annotation data.             

## Quick tour
```{r}
set.seed(123)   
library(richR)   
# To check if your the current species if supported !!!
showData()   
# Make the GO and KEGG Pathway data for your analysis
# find suitable species name by using showensemble()    
hsa_go<-buildAnnot(species="human",keytype="SYMBOL",anntype = "GO")
hsa_ko<-buildAnnot(species = "human",keytype="SYMBOL", anntype = "KEGG")
```   
### Support KEGG Module and MSIGDB anntation
```
hsamgi <- buildMSIGDB(species="human",keytype="SYMBOL",anntype="GO")
hsakom <- buildAnnot(species = "human",keytype="SYMBOL", anntype = "KEGGM")
```
____   

### You can make annotation data from ensemble   
* If had bioAnno installed, you can build annotation package with it  
```{r}
# library(bioAnno)
# fromKEGG(species="ath")
# athgo<-buildOwn(dbname="org.ath.eg.db",anntype="GO")  
# athko<-buildOwn(dbname="org.ath.eg.db",anntype="KEGG") 
# Please go over the bioAnno package webpage ("https://github.com/guokai8/bioAnno") to learn more
```   
----

### Simple example for enrichment analysis

```{r,fig.height=6,fig.width=6,fig.align="center",dpi=100}
gene <- sample(unique(hsa_go$GeneID),1000)
res <- richGO(gene,godata = hsa_go,ontology ="BP")
head(res)
ggbar(res,top = 20,usePadj = F)
```       

```{r,fig.height=6,fig.width=6,fig.align="center"}
resko<-richKEGG(gene,hsa_ko,pvalue=0.05)
head(resko)
ggdot(resko,top=10,usePadj = F)
```   
### Support DAVID analysis (Online)
```
gene <- sample(unique(hsako$GeneID),1000)
res <- richDAVID(gene,keytype="ENTREZID",species="human")
```
#### You can also get network graphic for any type of enrichment analysis result and also combine different enrichment result
```{r,fig.height=6,fig.width=6,fig.align="center",dpi=100}
ggnetplot(res,top=20)
ggnetwork(res,top=20,weightcut = 0.01)

```   
### Directly support dplyr filter, select, mutate,group_by ... functions
```
filter(resko,Padj<0.05)%>%head()
select(resko,Term)
```
### Generate figures with mutiple enrichment results for groups
```
gene1 <- sample(unique(hsako$GeneID),1000)
gene2 <- sample(unique(hsako$GeneID),1000)
resko1 <- richKEGG(gene1,kodata = hsako)
resko2 <- richKEGG(gene2,kodata = hsako)
res <- compareResult(list(S1=resko1,S2=resko2))
comparedot(res)
```
### Generate figures with different enrichment results
```
ggnetmap(list(res,resko),top=50,visNet=TRUE,smooth=FALSE)
```
### Contact information
For any questions please contact guokai8@gmail.com  

Congratulations ! You just done all Enrichment analysis 
