---
title: "richR: Functional Enrichment Analysis and Visualization"
author: "Kai Guo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{richR: Functional Enrichment Analysis and Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  dpi = 100
)
```

## Introduction

**richR** supports Over-Representation Analysis (ORA) via hypergeometric
testing, Gene Set Enrichment Analysis (GSEA), and kappa-score-based term
clustering. Built-in annotation builders cover GO, KEGG, Reactome, KEGG
Module, and MSigDB. You can also supply custom gene sets via GMT files or
named lists.

## Setup

```{r load-library}
set.seed(123)
library(richR)
has_orgdb <- requireNamespace("org.Hs.eg.db", quietly = TRUE)

# Check supported species
showData()
```

## 1. Build annotation data

```{r build-annot, eval=has_orgdb}
# GO and KEGG annotations for human
hsa_go <- buildAnnot(species = "human", keytype = "SYMBOL", anntype = "GO")
hsa_ko <- buildAnnot(species = "human", keytype = "SYMBOL", anntype = "KEGG")
```

KEGG Module and MSigDB annotations are also supported:

```r
hsakom <- buildAnnot(species = "human", keytype = "SYMBOL", anntype = "KEGGM")
hsamgi <- buildMSIGDB(species = "human", keytype = "SYMBOL", anntype = "GO")
```

### Custom gene sets

```r
# From GMT file (MSigDB, Enrichr, etc.)
annot <- readGMT("h.all.v2023.2.Hs.symbols.gmt", species = "human")
res   <- enrich(my_genes, annot)

# From a named list
my_sets <- list(
  "Apoptosis"  = c("TP53", "BAX", "BCL2", "CASP3"),
  "Cell Cycle" = c("CDK1", "CDK2", "CCND1", "RB1")
)
annot <- buildAnnotFromList(my_sets)
res   <- enrich(my_genes, annot)
```

## 2. Enrichment analysis

### GO enrichment (ORA)

```{r go-enrich, eval=has_orgdb}
gene  <- sample(unique(hsa_go$GeneID), 1000)
resgo <- richGO(gene, godata = hsa_go, ontology = "BP")
head(resgo)
```

### KEGG enrichment

```{r kegg-enrich, eval=has_orgdb}
resko <- richKEGG(gene, kodata = hsa_ko, pvalue = 0.05)
head(resko)
```

### GSEA

```{r gsea, eval=has_orgdb}
name     <- sample(unique(hsa_ko$GeneID), 1000)
genelist <- rnorm(1000)
names(genelist) <- name
res_gsea <- richGSEA(genelist, object = hsa_ko)
```

### DAVID (online)

```r
res_david <- richDAVID(gene, keytype = "ENTREZID", species = "human")
```

## 3. Visualization

### Bar and dot plots

```{r bar-dot, fig.height=6, eval=has_orgdb}
richBar(resgo, top = 20, usePadj = FALSE)
richDot(resko, top = 10, usePadj = FALSE)
```

### Lollipop plot

```{r lollipop, eval=has_orgdb}
richLollipop(resko, top = 20)
```

### Gene-level plots

```{r gene-plots, fig.height=6, fig.width=9, eval=has_orgdb}
# Gene-term dot plot (color by significance when no fold change provided)
richGeneDot(resgo)

# Gene-term heatmap with custom colors
richGeneHeat(resgo, usePadj = FALSE,
             low = "deepskyblue", high = "darkorange", mid = "white")

# Gene bar plot (x-axis = RichFactor, gene labels aligned left)
richGeneBar(resgo, top = 10)
```

### Network plots

```{r network, fig.height=6, eval=has_orgdb}
richNetplot(resko, top = 20)
richNetwork(resko, top = 20, weightcut = 0.01)
```

### GSEA-specific plots

```{r gsea-plots, eval=has_orgdb}
richNES(res_gsea, top = 20)
richVolcano(res_gsea)
```

## 4. Clustering and comparison

### Kappa-based clustering

```{r cluster, eval=has_orgdb}
resc <- richCluster(resgo)
richClusterDot(resc)
```

### Multi-group comparison

```{r compare, eval=has_orgdb}
gene1  <- sample(unique(hsa_ko$GeneID), 1000)
gene2  <- sample(unique(hsa_ko$GeneID), 1000)
resko1 <- richKEGG(gene1, kodata = hsa_ko)
resko2 <- richKEGG(gene2, kodata = hsa_ko)
res_cmp <- compareResult(list(S1 = resko1, S2 = resko2))
richCompareDot(res_cmp)
```

### Comparison heatmap

```{r heatmap, eval=has_orgdb}
richHeatmap(list(GO = resgo, KEGG = resko), top = 50)
```

### UpSet plot

```{r upset, fig.width=9, eval=has_orgdb}
gene_lists <- list(
  "Treatment A" = sample(unique(hsa_ko$GeneID), 500),
  "Treatment B" = sample(unique(hsa_ko$GeneID), 400),
  "Control"     = sample(unique(hsa_ko$GeneID), 600)
)
richUpset(gene_lists)
```

## 5. Working with results

```{r results, eval=has_orgdb}
# Extract result and detail tables
head(result(resgo))
head(detail(resgo))
```

### dplyr integration

```{r dplyr, eval=has_orgdb}
library(dplyr)
resgo %>% filter(Padj < 0.05) %>% head()
resko %>% select(Term, Pvalue, Padj) %>% head()
```

## Contact

For questions or bug reports please contact guokai8@gmail.com or open an
issue at <https://github.com/guokai8/richR/issues>.
