---
title: "richR: Functional Enrichment Analysis for GO and KEGG Pathways"
author: "Kai Guo, Junguk Hur"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{richR: Functional Enrichment Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette demonstrates how to use the richR package for functional enrichment
analysis of gene lists.

## Introduction

richR supports enrichment analysis for multiple annotation types including
GO (Gene Ontology), KEGG pathways, Reactome, and MSigDB gene sets. It provides
over-representation analysis (ORA) via hypergeometric testing, Gene Set Enrichment
Analysis (GSEA), and kappa-score-based term clustering, along with
publication-ready visualizations.

## Quick Start

```{r}
set.seed(123)
library(richR)

# Check supported species
showData()

# Build GO and KEGG annotation databases
hsa_go <- buildAnnot(species = "human", keytype = "SYMBOL", anntype = "GO")
hsa_ko <- buildAnnot(species = "human", keytype = "SYMBOL", anntype = "KEGG")
```

### MSigDB and KEGG Module Support

```
hsamsi <- buildMSIGDB(species = "human", keytype = "SYMBOL", anntype = "GO")
hsakom <- buildAnnot(species = "human", keytype = "SYMBOL", anntype = "KEGGM")
```

### Custom Annotation with bioAnno

If you have the [bioAnno](https://github.com/guokai8/bioAnno) package installed,
you can build annotation databases for additional species:

```{r}
# library(bioAnno)
# fromKEGG(species = "ath")
# athgo <- buildOwn(dbname = "org.ath.eg.db", anntype = "GO")
# athko <- buildOwn(dbname = "org.ath.eg.db", anntype = "KEGG")
```

## Over-Representation Analysis

### GO Enrichment

```{r, fig.height=6, fig.width=6, fig.align="center", dpi=100}
gene <- sample(unique(hsa_go$GeneID), 1000)
res <- richGO(gene, godata = hsa_go, ontology = "BP")
head(res)
ggbar(res, top = 20, usePadj = FALSE)

# Extract gene-term details
head(detail(res))
```

### Cluster GO Enrichment Terms

```{r}
resc <- richCluster(res)
ggdot(resc)
```

### KEGG Pathway Enrichment

```{r, fig.height=6, fig.width=6, fig.align="center"}
resko <- richKEGG(gene, hsa_ko, pvalue = 0.05)
head(resko)
ggdot(resko, top = 10, usePadj = FALSE)
```

## Gene Set Enrichment Analysis (GSEA)

```{r}
set.seed(123)
hsa_ko <- buildAnnot(species = "human", keytype = "SYMBOL", anntype = "KEGG")
name <- sample(unique(hsa_ko$GeneID), 1000)
gene <- rnorm(1000)
names(gene) <- name
res <- richGSEA(gene, object = hsa_ko)
```

## DAVID Analysis (Online)

Requires internet access and the `RDAVIDWebService` package:

```
gene <- sample(unique(hsa_ko$GeneID), 1000)
res <- richDAVID(gene, keytype = "ENTREZID", species = "human")
```

## Network Visualization

```{r, fig.height=6, fig.width=6, fig.align="center", dpi=100}
ggnetplot(resko, top = 20)
ggnetwork(res, top = 20, weightcut = 0.01)
```

## Tidyverse Integration

richR objects support `dplyr` verbs directly:

```
library(dplyr)
filter(resko, Padj < 0.05) %>% head()
select(resko, Term)
```

## Comparing Multiple Groups

```
gene1 <- sample(unique(hsa_ko$GeneID), 1000)
gene2 <- sample(unique(hsa_ko$GeneID), 1000)
resko1 <- richKEGG(gene1, kodata = hsa_ko)
resko2 <- richKEGG(gene2, kodata = hsa_ko)
res <- compareResult(list(S1 = resko1, S2 = resko2))
comparedot(res)
```

## Combined Network Visualization

```
ggnetmap(list(res, resko), top = 50, visNet = TRUE, smooth = FALSE)
```

## Contact

- Kai Guo: guokai8@gmail.com
- Junguk Hur: junguk.hur@med.und.edu
- Issues: https://github.com/hurlab/richR/issues
